% !TEX TS-program = lualatex
% !TeX spellcheck = en_GB
% !TeX root = ../My_thesis.tex

\addchap{Acknowledgement}
	\vspace{-2\baselineskip}
	I cannot thank Professor Andreas \"Ochsner enough; without his support and trust, this work could never have started and would have never finished. I am forever grateful for the opportunity of working under his supervision. I have learnt how a Professor could be indisputably knowledgeable and hard-working, yet down-to-earth and kind. I have received a tremendous amount of support continuously and particularly in times of need. I am also grateful for the dedication and support of Associate Professor Wayne Hall who mentored me in academia and reshaped my way of critical thinking. This work would have never been completed without his invaluable comments and worthwhile discussions. I am forever in his debt. I was fortunate enough to have two trusting principal supervisors, two wise mentors, and two patient friends at the same time. I will be taking all the acquired lessons throughout my life.
	
	I am thankful for the everlasting enthusiasm and support of Dr Nicholas Emerson. He joined at the final stage but his impression remains. I also want to thank Professor Hong Guan for initiating my Ph.D.~study at Griffith University. 
	
	I want to thank Griffith Graduate Research School for providing the initial financial support; and the academic, professional, and technical staff of the School of Engineering and Built Environment for the encouragement and support. I am thankful for the opportunity that was provided.
 
	Finally, I want to thank my parents and my brother without whom my existence would have been trivial.