% !TEX TS-program = lualatex
% !TeX spellcheck = en_GB
% !TeX root = ../My_thesis.tex

\addchap{List of Publications}
	The following contributions were produced to disseminate concepts and results from this thesis:
	\begin{itemize}
		\item Monograph
		\begin{itemize}
			\item Javanbakht, Z \& Öchsner, A (2017), Advanced finite element simulation with MSC Marc: Application of user subroutines, Cham: Springer.
		\end{itemize}
		\item Journal articles
	\begin{itemize}
		\item Javanbakht, Z, Hall, W \& Öchsner, A (2016), “Automatized Estimation of the Effective Thermal Conductivity of Carbon Fiber Reinforced Composite Materials”, Defect and Diffusion Forum 370, pp. 177–183.
		\item Javanbakht, Z, Hall, W \& Öchsner, A (2016), “Finite Element Evaluation of Effective Thermal Conductivity of Short Carbon Nano Tubes: A Comparative Study”, Defect and Diffusion Forum 372, pp. 208–214.
		\item Javanbakht, Z, Hall, W \& Öchsner, A (2017), “The Effect of Partitioning on the Clustering Index of Randomly-Oriented Fiber Composites: A Parametric Study”, Defect and Diffusion Forum 380, pp. 232–241.
		\item Javanbakht, Z, Hall, W, Virk, Amandeep Singh, Summerscales, John \& Öchsner, A (2019), ``Finite Element Analysis of Natural Fiber Composites Using a Self-Updating Model'', Journal of Composite Materials, Under review.
		\item Javanbakht, Z, Hall, W \& Öchsner, A (2019). “An Element-wise Scheme to Analyse Local Mechanical Anisotropy in Fibre-reinforced Composites”, Material Science and Technology, Under review.
	\end{itemize}
	\item Book chapters
	\begin{itemize}
		\item Javanbakht, Z, Hall, W \& Öchsner, A (2017), “Computational Evaluation of Transverse Thermal Conductivity of Natural Fiber Composites”, In: Improved Performance of Materials. Ed. by A Öchsner \& H Altenbach. Vol. 72. Advanced structured materials, Cham: Springer, pp. 197–206.
		\item Javanbakht, Z, Hall, W \& Öchsner, A (2019), “Effective Thermal Conductivity of Fiber Reinforced Composites Under Orientation Clustering”, In: Engineering design applications. Ed. by A Öchsner \& H Altenbach. Vol. 92. Advanced structured materials, 1869-8433, Cham, Switzerland: Springer, pp. 507–519.
	\end{itemize}
	\end{itemize}